### 2023-09-13 main.tex
asd
jll
ffg
1 to N

\vspace{.5cm}
\noindent
{\bf Remark:} in the following we will often consider type $1$ as a distinguished type
(called `holes'), whose number at each site is fixed by the numbers of particles 
of the remaining $N-1$ types (at the same site). It is interesting to observe that, in case we stop distinguishing the colors of the ``true'' particles of types $\{2,\ldots,N\}$, we retrieve the 
boundary-driven version of the partial symmetric exclusion process \cite{schutzSandow,carinci2013duality}.
The choice of type $1$ for the distinguished type is arbitrary and does not affect
the results. In particular, we shall see in Section \ref{sectionDuality} about duality, that
the dual process eventually has only particles of type $1$ on the graph.
Interpreting type $1$ as an hole, one can then say that the dual process voids the graph.

@@

The duality function is given by
\begin{equation}\label{dualityElements}
	D(\bm{n},\bm{\xi})=\prod_{x\in V}\left(\frac{(\nu -\sum_{a=2}^{N}\xi_{a}^{x})!}{\nu!}\prod_{a=2}^{N}\frac{n_{a}^{x}!}{(n_{a}^{x}-\xi_{a}^{x})!}\left(\rho_{a}^{x}\right)^{\xi_{a}^{u(x)}}\,\right)
\end{equation}
where we denote the \textit{density} of the species $a\in \{2,\ldots,N\}$ imposed by the reservoir at site $x\in V$  by 
\begin{equation}
\rho_{a}^{x}:=\frac{\alpha_{a}^{x}}{|\alpha^{x}|}
\end{equation}
The dual process is defined by its Markov generator, which reads
 \begin{equation}\label{DualGenerator}
    \widetilde{\mathcal{L}}=\sum_{(x,y)\in \mathcal{E}}\omega_{x,y}\mathcal{L}_{x,y}+\sum_{x\in V}\Gamma_{x}\widetilde{\mathcal{L}}_{x}
\end{equation}
where 
$\mathcal{L}_{x,y}$ is defined in \eqref{edgeGenerator} and, for any function $f:\widetilde{\Omega}\to \mathbb{R}$ 
\begin{equation}\label{siteDualGenerator}
    \widetilde{\mathcal{L}}_{x}f(\bm{\xi})=|\alpha^{x}|\sum_{a=2}^{N}\xi_{a}^{x}\left(f(\bm{\xi}-\bm{\delta}_{a}^{x}+\bm{\delta}_{1}^{x}+\bm{\delta}_{a}^{u(x)})-f(\bm{\xi})\right)
\end{equation}
\newline
Therefore the dual dynamics is described as follows. On one hand, the edge part $\mathcal{L}_{x,y}$ of the dual Markov generator gives rise to  the multi-species stirring dynamics on the graph. On the other hand, the site
part $\widetilde{\mathcal{L}}_{x}$ of the dual generator replaces a particle of any type $a\in\{2,\ldots,N\}$ at site $x$ with a particle of type $1$ and creates a particle of the same type $a$ at the extra-site $u(x)$. This last transition is performed with rate $|\alpha^{x}|\xi_{a}^{x}$. This means that eventually the dual process voids the graph, putting all the dual particles of species $\{2,\ldots,N\}$ in the extra-sites. In other words the extra-sites play the role of absorbing boundaries. 
\newline \newline
\textbf{Remark}: in the reversible situation, i.e. when $\forall x\in V$ $\rho_{a}^{x}=\rho_{a}$, the expectation  of the duality function  $D(\bm{n},\bm{\xi})$ with   $\bm{n}$ distributed as  $\mu_{rev} = \bigtimes_{x\in V}\text{Multinomial}\left(\twoj, \rho_{1},\ldots,\rho_{N}\right)$ is
\begin{equation}
\mathbb{E}_{\mu^{rev}}\left[D(\bm{n},\bm{\xi})\right]=\prod_{a=2}^{N}\left(\rho_{a}\right)^{\sum_{x\in V}\xi_{a}^{x}+\sum_{x\in V}\xi_{a}^{u(x)}}\qquad \forall \bm{n}\in \Omega,\quad\forall \bm{\xi}\in \widetilde{\Omega}
\end{equation}

\subsection{Proof of duality}
To prove duality between $(\bm{n}(t))_{t\geq 0}$ and $(\bm{\xi}(t))_{t\geq 0}$ we  show that \eqref{dualityIntertwines} is fulfilled.  To show this, we will use the Hamiltonians (linked with the Markov generators by \eqref{Hamiltonian-Generator}) and their Lie algebraic description. Indeed, in this formalism the proof reduces to finding symmetries and group like transformation of generators of the Lie algebra.\\
To a configuration $\bm{\xi}$ of the configuration space  \eqref{dualStateSpace} of a dual process we associate the vector
\begin{equation}
    |\bm{\xi}\rangle=\bigotimes_{x\in V}\left(|\xi_{1}^{x},\ldots,\xi_{N}^{x}\rangle\otimes |\xi_{1}^{u(x)},\ldots,\xi_{N}^{u(x)}\rangle\right)
\end{equation}
The Hamiltonian of the dual process reads
\begin{equation}\label{DualHamiltonian}
    \widetilde{H}=\sum_{x,y\in \mathcal{E}}\omega_{x,y}\mathcal{H}_{x,y}+\sum_{x\in V}\Gamma_{x}\widetilde{H}_{x}
\end{equation}
where $\mathcal{H}_{x,y}$ is the one defined in \eqref{edgeHamiltonian}, while 
\begin{equation}\label{siteDualHamiltonian}
    \widetilde{H}_{x}=|\alpha^{x}|\sum_{a=2}^{N}\left((a^{\dagger})_{a}^{u(x)}\,E_{1a}^{x}-E_{aa}^{x}\right)
\end{equation}
Here we introduced bosonic creation operator $a^{\dagger}$ acting on the left-hand-side as $\langle q|a^{\dagger}=\langle q+1|$ on a generic vector $\langle q|$ with $q\in \mathbb{N}_{0}$, so that in \eqref{siteDualHamiltonian} 
$(a^{\dagger})_{a}^{u(x)}$ denotes a copy of $a^{\dagger}$ acting on the extra-site $u(x)$ and on the species $a\in\{2,\ldots,N\}$. \\
We will show below that the Hamiltonians \eqref{OriginalHamiltonian} and \eqref{DualHamiltonian} are dual in the sense of \eqref{DualityRelation}. The duality matrix $D$ is described as 
\begin{equation}\label{dualityMatrix}
    D=\prod_{x\in V}d_{x}\otimes \dd
\end{equation}
where
\begin{equation}\label{bulkElementDualityMatrix}
d_{x}=R_{x}\exp{(E^{x})}
\end{equation}
with 
\begin{equation}\label{EquationEx}
E^{x}=\sum_{a=2}^{N}E_{a1}^{x}
\end{equation}
and
\begin{equation}\label{Revmatrix}
    R_{x}=\sum_{n^{x}\in\Omega_{x}}\frac{\prod_{a=1}^{N}n_{a}^{x}!}{\nu!}|n_{1}^{x},\ldots,n_{N}^{x}\rangle\langle n_{1}^{x},\ldots,n_{N}^{x}|
\end{equation}
and where 
\begin{equation}\label{dualityMatrix2}
\dd=\sum_{\xi_{2}^{u(x)},\ldots,\xi_{N}^{u(x)}=0}^{\infty}\prod_{a=2}^{N}\left(\rho_{a}^{x}\right)^{\xi_{a}^{u(x)}}\langle \xi_{1}^{u(x)},\ldots,\xi_{N}^{u(x)}|
\end{equation}
The matrix $R_{x}$ is diagonal. Its elements are related to the inverse of the weights of the reversible measure \eqref{reversibleMeasure}. In particular, to obtain these elements, we have considered the weights of \eqref{reversibleMeasure} when all the parameters $p_{i}=\frac{1}{N}$. Then, the constant $\left(\frac{1}{N}\right)^{\nu}$ has been neglected, since it does not change the duality relation. This $R_{x}$ is called the ``cheap'' duality matrix (see \cite{giardina2009duality}). \\
Since \eqref{dualityMatrix} is product over sites, proving \eqref{DualityRelation} is equivalent to showing that 
\begin{equation}\label{edgeDualRealtion}
    \mathcal{H}_{x,y}^{T}D=D\mathcal{H}_{x,y}\qquad \forall (x,y)\in \mathcal{E}
\end{equation}
and 
\begin{equation}\label{siteDualRelation}
    H_{x}^{T}D=D\widetilde{H}_{x}\qquad \forall x\in V.
\end{equation}
We perform the proof of duality in three steps: first we will show that matrix \eqref{dualityMatrix} has elements \eqref{dualityElements}; second we will prove \eqref{edgeDualRealtion}; finally we will show \eqref{siteDualRelation}. 
As a preliminary result we note that 
\begin{equation}\label{transpositionPropertyR}
(E_{ba}^{x})^{T}=R_{x}E_{ab}^{x}R_{x}^{-1}\qquad \forall x\in V
\end{equation}
that follows immediately from the definition of $E_{ab}$. 
\begin{comment}

@@

	 &\langle n^{x}|d_{x}\otimes \dd|\xi^{x}\rangle\\=&\langle n_{1}^{x},\ldots,n_{N}^{x}| (\exp{(E_{12}^{x}+\ldots+E_{1N}^{x}}))^{T}R_{x}\otimes\sum_{q_{2}^{u(x)},\ldots,q_{N}^{u(x)}=0}^{\infty}\prod_{a=2}^{N}\left(\rho_{a}^{x}\right)^{q_{a}^{u(x)}}\langle q_{1}^{u(x)},\ldots,q_{N}^{u(x)}|

@@

\sum_{q_{2}^{u(x)},\ldots,q_{N}^{u(x)}=0}^{\infty}\prod_{a=2}^{N}\left(\rho_{a}^{x}\right)^{q_{a}^{u(x)}}\langle q_{1}^{u(x)},\ldots,q_{N}^{u(x)}|\xi_{1}^{u(x)},\ldots,\xi_{N}^{u(x)}\rangle=\prod_{a=2}^{N}\left(\rho_{a}^{x}\right)^{\xi_{a}^{u(x)}}

@@

&\langle n_{1}^{x},\ldots,n_{N}^{x}|(\exp{(E_{12}^{x}+\ldots+E_{1N}^{x})})^{T}R_{x}|\xi_{1}^{x},\ldots,\xi_{N}^{x}\rangle\\&= \langle  n_{1}^{x},\ldots,n_{N}^{x}|\left(\sum_{k_{2}=0}^{\infty}\frac{\left\{\left(E_{12}^{x}\right)^{T}\right\}^{k_{2}}}{k_{2}!}\ldots\sum_{k_{N}=0}^{\infty}\frac{\left\{\left(E_{1N}^{x}\right)^{T}\right\}^{k_{N}}}{k_{N}!}\sum_{s\in\Omega_{x}}\frac{s_{1}^{x}!\ldots s_{N}^{x}!}{\nu!}|s_{1}^{x},\ldots,s_{N}^{x}\rangle\langle s_{1}^{x},\ldots,s_{N}^{x}|\right)

@@

\sum_{k_{2}=0}^{n_{1}^{x}}\ldots\sum_{k_{N}=0}^{n_{N}^{x}}\langle n_{1}^{x}+k_{2}+\ldots+k_{N},n_{2}^{x}-k_{2},\ldots,n_{N}^{x}-k_{N}|

@@

\hspace{2.cm}\prod_{a=2}^{N}\left((n_{a}^{x}(n_{a}^{x}-1)\ldots (n_{a}^{x}-k_{a}+1)\right)\frac{1}{k_{2}!,\ldots,k_{N}!}\frac{\xi_{1}^{x}!\ldots \xi_{N}^{x}!}{\nu!}|\xi_{1}^{x},\ldots,\xi_{N}^{x}\rangle
\sum_{k_{2}=0}^{n_{1}^{x}}\ldots\sum_{k_{N}=0}^{n_{N}^{x}}\langle n_{1}^{x}+k_{2}+\ldots+k_{N},n_{2}^{x}-k_{2},\ldots,n_{N}^{x}-k_{N}|\frac{n_{2}^{x}!\ldots n_{N}^{x}!}{(n_{2}^{x}-k_{2})!\ldots(n_{N}^{x}-k_{N})!}

@@

\hspace{2.cm}\qquad\frac{1}{k_{2}!,\ldots,k_{N}!}\frac{\xi_{1}^{x}!\ldots \xi_{N}^{x}!}{\nu!}|\xi_{1}^{x},\ldots,\xi_{N}^{x}\rangle
\frac{(\nu-\sum_{a=2}^{N}\xi_{a}^{x})}{\nu!}\prod_{a=2}^{N}\frac{n_{a}^{x}!}{(n_{a}^{x}-\xi_{a}^{x})!}

@@

where in the last equality we applied the orthogonality relations \eqref{ortho} and the fact that $\xi_{1}^{x}=\nu-\sum_{a=2}^{N}\xi_{a}^{x}$. Finally, by taking the product over $x\in V$ \eqref{proofDualityElements} is proved.

@@

\paragraph{Proof of \eqref{edgeDualRealtion}.}To show this relation we need two 'ingredients'. First the existence of a similarity transformation between the Hamiltonian $\mathcal{H}_{x,y}$  and its transposed. As we will show, this similarity transformation is $R_{x}R_{y}$.  Second, the possibility of finding a symmetry for the edge Hamiltonian. Exploiting \eqref{hamiltonianCasimir}, we will search for a symmetry of the second Casimir $C_{2}$. As we will see, this symmetry is $\sum_{a=2}^{N}E_{a1}$ \\

@@

\begin{equation}\label{transpositionPropertyH}
    \mathcal{H}_{x,y}^{T}=\left(R_{x}R_{y}\right)\mathcal{H}_{x,y}\left(R_{x}R_{y}\right)^{-1}
\end{equation}
therefore we have found the similarity transformation between $\mathcal{H}_{x,y}$ and its transposed. \\
We now look for a symmetry of $\mathcal{H}_{x,y}$, i.e. a matrix $S_{x,y}$ of the same dimension such that it satisfies 
\begin{equation}
	\mathcal{H}_{x,y}S_{x,y}=S_{x,y}\mathcal{H}_{x,y}\quad \Leftrightarrow \quad [\mathcal{H}_{x,y},S_{x,y}]=0
\end{equation}
Using \eqref{hamiltonianCasimir}, we observe that $\mathcal{H}_{x,y}$ is proportional to the coproduct of the second Casimir, up to a diagonal term. Therefore, it is enough to look for a symmetry of $\Delta (C_{2})$. Using the bilinearity of the coproduct operator and the fact that $C_{2}$ belongs to the centre of the Lie algebra, it is easy to show that for any matrix $E\in gl(N)$ it holds
\begin{equation}
	[C_{2},E]=0\quad \Rightarrow\quad \left[\Delta (C_{2}),\Delta(E)) \right]=0
\end{equation}
We chose 
\begin{equation}
	E=\sum_{a=2}^{N}E_{a1}
\end{equation}
and, for a fixed $(x,y)\in \mathcal{E}$, we define
\begin{equation}
	S_{x,y}=\exp{(\Delta^{x,y}(E))}=\exp{(E^{x})}\exp{(E^{y})}
\end{equation}
where $\Delta^{x,y}(E)$ is a copy of the coproduct acting on sites $x$ and $y$. Therefore, this operator $S_{x,y}$ satisfies
\begin{equation}\label{symmetryH}
	\left[S_{x,y},\mathcal{H}_{x,y}\right]=0
\end{equation}
i.e. it is a symmetry of $\mathcal{H}_{x,y}$. \\ Exploiting these considerations we write

@@

% However, they would lead to duality matrices where the number of dual particles is greater than the number of original ones, loosing the simplification of the dynamics of the dual process. 
 \begin{flushright}
     $\square$
 \end{flushright}
 \paragraph{Proof of \eqref{siteDualRelation}.} To prove \eqref{siteDualRelation} we need to transform via the Hadamard formula the transposed of the site Hamiltonian \eqref{siteDualHamiltonian} and then to  introduce properly a creation operator acting on an extra-site $u(x)$. % that we connect to every site $x\in V$ of the graph $G$. \\
 Considering $A,B\in gl(N)$, the Hadamard formula reads 
 \begin{equation}\label{HadamardFormula}
     \exp{(-B)}A\exp{(B)}=A-\left[B,A\right]+\frac{1}{2!}\left[B,\left[B,A\right]\right]-\frac{1}{3!}\left[B,\left[B,\left[B,A\right]\right]\right]+\ldots
 \end{equation}
In the following we need the application of the Hadamard formula with $B=\sum_{a=2}^{N}E_{a1}$ defined in \eqref{equationE} and with $A$ equal to some of the generator $E_{ab}$ of the Lie algebra. Therefore we compute 
\begin{enumerate}
    \item for $A=E_{b 1}$ with $b\in \{2,\ldots,N\}$, we obtain 
    \begin{equation}\label{HT_El1}
        \exp{(-E)}E_{b 1}\exp{(E)}=E_{b 1}
    \end{equation}
    because $E_{b 1}$ commutes with $E$
    \item for $A=E_{bb}$ with $b \in \{2,\ldots,N\}$, we obtain 
    \begin{equation}\label{HT_Ell}
        \exp{(-E)}E_{b b}\exp{(E)}=E_{b b}+E_{b 1}
    \end{equation}
Indeed, using \eqref{eq:comgl} we have 
    \begin{equation}
       \left[E, E_{bb}\right]=\sum_{a=2}^{N}\left(E_{ab}\delta_{b 1}-E_{b 1}\delta_{ab}\right)=-E_{b1}
    \end{equation}
    Inserting the above commutator in \eqref{HadamardFormula}, we obtain \eqref{HT_Ell}.
    \item for $A=E_{1b}$ with $b\in \{2,\ldots,N\}$, we obtain 
    \begin{equation}\label{HT-E1l}
        \exp{(-E)}E_{1 a}\exp{(E)}=E_{1a}+E_{11}-\sum_{c=2}^{N}\left(E_{c1}+E_{ca}\right)
    \end{equation}
  %\com{Why indices have names $b, k, j, a, b$? Fix one set for the generators}

>>>>

\vspace{.5cm}
\noindent
{\bf Remark:} in the following we will often consider type $N$ as a distinguished type
(called `holes'), whose number at each site is fixed by the numbers of particles 
of the remaining $N-1$ types (at the same site). It is interesting to observe that, in case we stop distinguishing the colors of the ``true'' particles of types $\{1,\ldots,N-1\}$, we retrieve the 
boundary-driven version of the partial symmetric exclusion process \cite{schutzSandow,carinci2013duality}.
The choice of type $N$ for the distinguished type is arbitrary and does not affect
the results. In particular, we shall see in Section \ref{sectionDuality} about duality, that
the dual process eventually has only particles of type $N$ on the graph.
Interpreting type $N$ as an hole, one can then say that the dual process voids the graph.

@@

The duality function is given by
\begin{equation}\label{dualityElements}
	D(\bm{n},\bm{\xi})=\prod_{x\in V}\left(\frac{(\nu -\sum_{a=2}^{N}\xi_{a}^{x})!}{\nu!}\prod_{a=1}^{N-1}\frac{n_{a}^{x}!}{(n_{a}^{x}-\xi_{a}^{x})!}\left(\rho_{a}^{x}\right)^{\xi_{a}^{u(x)}}\,\right)
\end{equation}
where we denote the \textit{density} of the species $a\in \{2,\ldots,N\}$ imposed by the reservoir at site $x\in V$  by 
\begin{equation}
\rho_{a}^{x}:=\frac{\alpha_{a}^{x}}{|\alpha^{x}|}
\end{equation}
The dual process is defined by its Markov generator, which reads
 \begin{equation}\label{DualGenerator}
    \widetilde{\mathcal{L}}=\sum_{(x,y)\in \mathcal{E}}\omega_{x,y}\mathcal{L}_{x,y}+\sum_{x\in V}\Gamma_{x}\widetilde{\mathcal{L}}_{x}
\end{equation}
where 
$\mathcal{L}_{x,y}$ is defined in \eqref{edgeGenerator} and, for any function $f:\widetilde{\Omega}\to \mathbb{R}$ 
\begin{equation}\label{siteDualGenerator}
    \widetilde{\mathcal{L}}_{x}f(\bm{\xi})=|\alpha^{x}|\sum_{a=1}^{N-1}\xi_{a}^{x}\left(f(\bm{\xi}-\bm{\delta}_{a}^{x}+\bm{\delta}_{1}^{x}+\bm{\delta}_{a}^{u(x)})-f(\bm{\xi})\right)
\end{equation}
\newline
Therefore the dual dynamics is described as follows. On one hand, the edge part $\mathcal{L}_{x,y}$ of the dual Markov generator gives rise to  the multi-species stirring dynamics on the graph. On the other hand, the site
part $\widetilde{\mathcal{L}}_{x}$ of the dual generator replaces a particle of any type $a\in\{2,\ldots,N\}$ at site $x$ with a particle of type $N$ and creates a particle of the same type $a$ at the extra-site $u(x)$. This last transition is performed with rate $|\alpha^{x}|\xi_{a}^{x}$. This means that eventually the dual process voids the graph, putting all the dual particles of species $\{1,\ldots,N-1\}$ in the extra-sites. In other words the extra-sites play the role of absorbing boundaries. 
\newline \newline
\textbf{Remark}: in the reversible situation, i.e. when $\forall x\in V$ $\rho_{a}^{x}=\rho_{a}$, the expectation  of the duality function  $D(\bm{n},\bm{\xi})$ with   $\bm{n}$ distributed as  $\mu_{rev} = \bigtimes_{x\in V}\text{Multinomial}\left(\twoj, \rho_{1},\ldots,\rho_{N}\right)$ is
\begin{equation}
\mathbb{E}_{\mu^{rev}}\left[D(\bm{n},\bm{\xi})\right]=\prod_{a=1}^{N-1}\left(\rho_{a}\right)^{\sum_{x\in V}\xi_{a}^{x}+\sum_{x\in V}\xi_{a}^{u(x)}}\qquad \forall \bm{n}\in \Omega,\quad\forall \bm{\xi}\in \widetilde{\Omega}
\end{equation}

\subsection{Proof of duality}
To prove duality between $(\bm{n}(t))_{t\geq 0}$ and $(\bm{\xi}(t))_{t\geq 0}$ we  show that \eqref{dualityIntertwines} is fulfilled.  To show this, we will use the Hamiltonians (linked with the Markov generators by \eqref{Hamiltonian-Generator}) and their Lie algebraic description. Indeed, in this formalism the proof reduces to finding symmetries and group like transformation of generators of the Lie algebra.\\
To a configuration $\bm{\xi}$ of the configuration space  \eqref{dualStateSpace} of a dual process we associate the vector
\begin{equation}
    |\bm{\xi}\rangle=\bigotimes_{x\in V}\left(|\xi_{1}^{x},\ldots,\xi_{N}^{x}\rangle\otimes |\xi_{1}^{u(x)},\ldots,\xi_{N}^{u(x)}\rangle\right)
\end{equation}
The Hamiltonian of the dual process reads
\begin{equation}\label{DualHamiltonian}
    \widetilde{H}=\sum_{x,y\in \mathcal{E}}\omega_{x,y}\mathcal{H}_{x,y}+\sum_{x\in V}\Gamma_{x}\widetilde{H}_{x}
\end{equation}
where $\mathcal{H}_{x,y}$ is the one defined in \eqref{edgeHamiltonian}, while 
\begin{equation}\label{siteDualHamiltonian}
    \widetilde{H}_{x}=|\alpha^{x}|\sum_{a=1}^{N-1}\left((a^{\dagger})_{a}^{u(x)}\,E_{Na}^{x}-E_{aa}^{x}\right)
\end{equation}
Here we introduced bosonic creation operator $a^{\dagger}$ acting on the left-hand-side as $\langle q|a^{\dagger}=\langle q+1|$ on a generic vector $\langle q|$ with $q\in \mathbb{N}_{0}$, so that in \eqref{siteDualHamiltonian} 
$(a^{\dagger})_{a}^{u(x)}$ denotes a copy of $a^{\dagger}$ acting on the extra-site $u(x)$ and on the species $a\in\{1,\ldots,N-1\}$. \\
We will show below that the Hamiltonians \eqref{OriginalHamiltonian} and \eqref{DualHamiltonian} are dual in the sense of \eqref{DualityRelation}. The duality matrix $D$ is described as 
\begin{equation}\label{dualityMatrix}
    D=\prod_{x\in V}d_{x}\otimes \dd
\end{equation}
where
\begin{equation}\label{bulkElementDualityMatrix}
d_{x}=R_{x}\exp{(E^{x})}
\end{equation}
with 
\begin{equation}\label{EquationEx}
E^{x}=\sum_{a=1}^{N-1}E_{aN}^{x}
\end{equation}
and
\begin{equation}\label{Revmatrix}
    R_{x}=\sum_{n^{x}\in\Omega_{x}}\frac{\prod_{a=1}^{N}n_{a}^{x}!}{\nu!}|n_{1}^{x},\ldots,n_{N}^{x}\rangle\langle n_{1}^{x},\ldots,n_{N}^{x}|
\end{equation}
and where 
\begin{equation}\label{dualityMatrix2}
\dd=\sum_{\xi_{1}^{u(x)},\ldots,\xi_{N-1}^{u(x)}=0}^{\infty}\prod_{a=1}^{N-1}\left(\rho_{a}^{x}\right)^{\xi_{a}^{u(x)}}\langle \xi_{1}^{u(x)},\ldots,\xi_{N}^{u(x)}|
\end{equation}
The matrix $R_{x}$ is diagonal. Its elements are related to the inverse of the weights of the reversible measure \eqref{reversibleMeasure}. In particular, to obtain these elements, we have considered the weights of \eqref{reversibleMeasure} when all the parameters $p_{i}=\frac{1}{N}$. Then, the constant $\left(\frac{1}{N}\right)^{\nu}$ has been neglected, since it does not change the duality relation. This $R_{x}$ is called the ``cheap'' duality matrix (see \cite{giardina2009duality}). \\
Since \eqref{dualityMatrix} is product over sites, proving \eqref{DualityRelation} is equivalent to showing that 
\begin{equation}\label{edgeDualRealtion}
    \mathcal{H}_{x,y}^{T}D=D\mathcal{H}_{x,y}\qquad \forall (x,y)\in \mathcal{E}
\end{equation}
and 
\begin{equation}\label{siteDualRelation}
    H_{x}^{T}D=D\widetilde{H}_{x}\qquad \forall x\in V.
\end{equation}
We perform the proof of duality in three steps: first we will show that matrix \eqref{dualityMatrix} has elements \eqref{dualityElements}; second we will prove \eqref{edgeDualRealtion}; finally we will show \eqref{siteDualRelation}. 
As a preliminary result we note that 
\begin{equation}\label{transpositionPropertyR}
(E_{ba}^{x})^{T}=R_{x}E_{ab}^{x}R_{x}^{-1}\qquad \forall x\in V
\end{equation}
that follows immediately from the definition of $E_{ab}$. 
\begin{comment}

@@

	 &\langle n^{x}|d_{x}\otimes \dd|\xi^{x}\rangle\\=&\langle n_{1}^{x},\ldots,n_{N}^{x}| (\exp{(E_{N1}^{x}+\ldots+E_{N\,N-1}^{x}}))^{T}R_{x}\otimes\sum_{q_{1}^{u(x)},\ldots,q_{N-1}^{u(x)}=0}^{\infty}\prod_{a=1}^{N-1}\left(\rho_{a}^{x}\right)^{q_{a}^{u(x)}}\langle q_{1}^{u(x)},\ldots,q_{N}^{u(x)}|

@@

\sum_{q_{1}^{u(x)},\ldots,q_{N-1}^{u(x)}=0}^{\infty}\prod_{a=1}^{N-1}\left(\rho_{a}^{x}\right)^{q_{a}^{u(x)}}\langle q_{1}^{u(x)},\ldots,q_{N}^{u(x)}|\xi_{1}^{u(x)},\ldots,\xi_{N}^{u(x)}\rangle=\prod_{a=1}^{N-1}\left(\rho_{a}^{x}\right)^{\xi_{a}^{u(x)}}

@@

&\langle n_{1}^{x},\ldots,n_{N}^{x}|(\exp{(E_{N1}^{x}+\ldots+E_{N\,N-1}^{x})})^{T}R_{x}|\xi_{1}^{x},\ldots,\xi_{N}^{x}\rangle\\&= \langle  n_{1}^{x},\ldots,n_{N}^{x}|\left(\sum_{k_{1}=0}^{\infty}\frac{\left\{\left(E_{N1}^{x}\right)^{T}\right\}^{k_{1}}}{k_{1}!}\ldots\sum_{k_{N-1}=0}^{\infty}\frac{\left\{\left(E_{N\,N-1}^{x}\right)^{T}\right\}^{k_{N-1}}}{k_{N-1}!}\sum_{s\in\Omega_{x}}\frac{s_{1}^{x}!\ldots s_{N}^{x}!}{\nu!}|s_{1}^{x},\ldots,s_{N}^{x}\rangle\langle s_{1}^{x},\ldots,s_{N}^{x}|\right)

@@

\sum_{k_{1}=0}^{n_{1}^{x}}\ldots\sum_{k_{N-1}=0}^{n_{N-1}^{x}}\langle n_{1}^{x}-k_{1},\ldots,n_{N-1}^{x}-k_{N-1},n_{N}^{x}+k_{1}+\ldots+k_{N-1}|

@@

\hspace{2.cm}\prod_{a=1}^{N-1}\left((n_{a}^{x}(n_{a}^{x}-1)\ldots (n_{a}^{x}-k_{a}+1)\right)\frac{1}{k_{1}!,\ldots,k_{N-1}!}\frac{\xi_{1}^{x}!\ldots \xi_{N}^{x}!}{\nu!}|\xi_{1}^{x},\ldots,\xi_{N}^{x}\rangle
\sum_{k_{1}=0}^{n_{1}^{x}}\ldots\sum_{k_{N-1}=0}^{n_{N-1}^{x}}\langle n_{1}^{x}-k_{1},\ldots,n_{N-1}^{x}-k_{N-1},\ldots,n_{N}^{x}+k_{1}+\ldots+k_{N-1}|\frac{n_{1}^{x}!\ldots n_{N-1}^{x}!}{(n_{1}^{x}-k_{1})!\ldots(n_{N-1}^{x}-k_{N-1})!}

@@

\hspace{2.cm}\qquad\frac{1}{k_{1}!,\ldots,k_{N-1}!}\frac{\xi_{1}^{x}!\ldots \xi_{N}^{x}!}{\nu!}|\xi_{1}^{x},\ldots,\xi_{N}^{x}\rangle
\frac{(\nu-\sum_{a=1}^{N-1}\xi_{a}^{x})}{\nu!}\prod_{a=1}^{N-1}\frac{n_{a}^{x}!}{(n_{a}^{x}-\xi_{a}^{x})!}

@@

where in the last equality we applied the orthogonality relations \eqref{ortho} and the fact that $\xi_{N}^{x}=\nu-\sum_{a=1}^{N-1}\xi_{a}^{x}$. Finally, by taking the product over $x\in V$ \eqref{proofDualityElements} is proved.

@@

\paragraph{Proof of \eqref{edgeDualRealtion}.}To show this relation we need two 'ingredients'. First the existence of a similarity transformation between the Hamiltonian $\mathcal{H}_{x,y}$  and its transposed. As we will show, this similarity transformation is $R_{x}R_{y}$.  Second, the possibility of finding a symmetry for the edge Hamiltonian. Exploiting \eqref{hamiltonianCasimir}, we will search for a symmetry of the second Casimir $C_{2}$. As we will see, this symmetry is $\sum_{a=1}^{N-1}E_{a1}$ \\

@@

\begin{equation}\label{transpositionPropertyH}
    \mathcal{H}_{x,y}^{T}=\left(R_{x}R_{y}\right)\mathcal{H}_{x,y}\left(R_{x}R_{y}\right)^{-1}
\end{equation}
therefore we have found the similarity transformation between $\mathcal{H}_{x,y}$ and its transposed. \\
We now look for a symmetry of $\mathcal{H}_{x,y}$, i.e. a matrix $S_{x,y}$ of the same dimension such that it satisfies 
\begin{equation}
	\mathcal{H}_{x,y}S_{x,y}=S_{x,y}\mathcal{H}_{x,y}\quad \Leftrightarrow \quad [\mathcal{H}_{x,y},S_{x,y}]=0
\end{equation}
Using \eqref{hamiltonianCasimir}, we observe that $\mathcal{H}_{x,y}$ is proportional to the coproduct of the second Casimir, up to a diagonal term. Therefore, it is enough to look for a symmetry of $\Delta (C_{2})$. Using the bilinearity of the coproduct operator and the fact that $C_{2}$ belongs to the centre of the Lie algebra, it is easy to show that for any matrix $E\in gl(N)$ it holds
\begin{equation}
	[C_{2},E]=0\quad \Rightarrow\quad \left[\Delta (C_{2}),\Delta(E)) \right]=0
\end{equation}
We chose 
\begin{equation}
	E=\sum_{a=1}^{N-1}E_{a1}
\end{equation}
and, for a fixed $(x,y)\in \mathcal{E}$, we define
\begin{equation}
	S_{x,y}=\exp{(\Delta^{x,y}(E))}=\exp{(E^{x})}\exp{(E^{y})}
\end{equation}
where $\Delta^{x,y}(E)$ is a copy of the coproduct acting on sites $x$ and $y$. Therefore, this operator $S_{x,y}$ satisfies
\begin{equation}\label{symmetryH}
	\left[S_{x,y},\mathcal{H}_{x,y}\right]=0
\end{equation}
i.e. it is a symmetry of $\mathcal{H}_{x,y}$. \\ Exploiting these considerations we write

@@

% However, they would lead to duality matrices where the number of dual particles is greater than the number of original ones, loosing the simplification of the dynamics of the dual process. 
 \begin{flushright}
     $\square$
 \end{flushright}
 \paragraph{Proof of \eqref{siteDualRelation}.} To prove \eqref{siteDualRelation} we need to transform via the Hadamard formula the transposed of the site Hamiltonian \eqref{siteDualHamiltonian} and then to  introduce properly a creation operator acting on an extra-site $u(x)$. % that we connect to every site $x\in V$ of the graph $G$. \\
 Considering $A,B\in gl(N)$, the Hadamard formula reads 
 \begin{equation}\label{HadamardFormula}
     \exp{(-B)}A\exp{(B)}=A-\left[B,A\right]+\frac{1}{2!}\left[B,\left[B,A\right]\right]-\frac{1}{3!}\left[B,\left[B,\left[B,A\right]\right]\right]+\ldots
 \end{equation}
In the following we need the application of the Hadamard formula with $B=\sum_{a=1}^{N-1}E_{a1}$ defined in \eqref{equationE} and with $A$ equal to some of the generator $E_{ab}$ of the Lie algebra. Therefore we compute 
\begin{enumerate}
    \item for $A=E_{b N}$ with $b\in \{1,\ldots,N-1\}$, we obtain 
    \begin{equation}\label{HT_El1}
        \exp{(-E)}E_{b N}\exp{(E)}=E_{b N}
    \end{equation}
    because $E_{b 1}$ commutes with $E$
    \item for $A=E_{bb}$ with $b \in \{1,\ldots,N-1\}$, we obtain 
    \begin{equation}\label{HT_Ell}
        \exp{(-E)}E_{b b}\exp{(E)}=E_{b b}+E_{b N}
    \end{equation}
Indeed, using \eqref{eq:comgl} we have 
    \begin{equation}
       \left[E, E_{bb}\right]=\sum_{a=1}^{N-1}\left(E_{ab}\delta_{b N}-E_{b N}\delta_{ab}\right)=-E_{bN}
    \end{equation}
    Inserting the above commutator in \eqref{HadamardFormula}, we obtain \eqref{HT_Ell}.
    \item for $A=E_{Nb}$ with $b\in \{1,\ldots,N-1\}$, we obtain 
    \begin{equation}\label{HT-E1l}
        \exp{(-E)}E_{N a}\exp{(E)}=E_{Na}+E_{NN}-\sum_{c=1}^{N-1}\left(E_{cN}+E_{ca}\right)
    \end{equation}
  %\com{Why indices have names $b, k, j, a, b$? Fix one set for the generators}
